A pipeline to analyse time-course gene expression data

### 2019-08-26 scripts/manuscript.Rmd
Renaming epicon lfc into timecourse

| Name   | Formula   |     |
|------- | --------- |--- |
| timely | $\text{lfc}(t)$  | Function of time |
| sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
| abs_sum  | $\sum_t \|lfc(t)\|$  | Always positive |
| max  | $\max_t \|lfc(t)\|$  | Always positive  |
| min  | $\min_t \|lfc(t)\|$  | Always positive |
| epicon  | $$  | Captures overall strength of response and overall direction |

**EAP: `epicon` is going to be weird for general use**

```{r}
log_fold_change_epicon = estimate_log_fold_change(
    data, splines_model, timecourse_contrasts,  method="epicon")

@@

```{r echo=FALSE, fig.height=6, fig.width=6, message=FALSE}
par(mfrow=c(2, 2))
plot(log_fold_change_epicon[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="EPICON vs sum",
     xlab="EPICON", ylab="sum")
plot(log_fold_change_epicon[, "M-C"],
     log_fold_change_max[, "M-C"], pch=20, main="EPICON vs max",
     xlab="EPICON", ylab="max")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
     xlab="max", ylab="min")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
     xlab="max", ylab="sum")
```

From the single measures of log-fold change and the p-value, we can now look
at the volcano plot. Here is an example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the EPICON log fold change
computation.

```{r fig.height=4, fig.width=4}
pvalue = qvalues[, "M-K"]
lfc_epicon = log_fold_change_epicon[, "M-K"]
names(lfc_epicon) = row.names(log_fold_change_epicon)

plot(lfc_epicon, -log10(pvalue), pch=20, main="Volcano plot",
     xlim=c(-2.5, 2))
```

@@

And here we visualize the genes with the largest absolute EPICON log-fold
change.

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_epicon),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

# convert gene names
genes = getBM(attributes=c("ensembl_gene_id", "entrezgene_id"),
		filters="refseq_mrna", values=genes,
		mart=ensembl)["entrezgene_id"]
genes = as.vector(unlist(genes))
pathways = find_enriched_pathway(
    genes, species="mmu",
    download_latest=TRUE)$stastic

>>>>

| Name   | Formula   |     |
|------- | --------- |--- |
| timely | $\text{lfc}(t)$  | Function of time |
| sum  | $\sum_t lfc(t)$  | Sum of log fold change. |
| abs_sum  | $\sum_t \lvert lfc(t)\lvert$  | Always positive |
| max  | $\max_t \lvert lfc(t) \lvert$  | Always positive  |
| min  | $\min_t \lvert lfc(t) \lvert$  | Always positive |
| timecourse | See details below | Captures overall strength and direction  |

To capture the overall strength and direction of the response, we leverage the
timepoint by timepoint log-fold change $\text{lfc}(t)$, and apply the
following formula:

$$\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T lfc(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {lfc(t)} \lvert \right)$$

Note that when a gene is not consistently up or down-regulated the estimation
of the direction will not represent accurately the changes observed.

```{r}
log_fold_change_timecourse = estimate_log_fold_change(
    data, splines_model, timecourse_contrasts,  method="timecourse")

@@

```{r echo=FALSE, fig.height=6, fig.width=6, message=FALSE}
par(mfrow=c(2, 2))
plot(log_fold_change_timecourse[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="timecourse vs sum",
     xlab="timecourse", ylab="sum")
plot(log_fold_change_timecourse[, "M-C"],
     log_fold_change_max[, "M-C"], pch=20, main="timecourse vs max",
     xlab="timecourse", ylab="max")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
     xlab="max", ylab="min")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
     xlab="max", ylab="sum")
```

From the single measures of log-fold change and the p-value, we can now look
at the volcano plot. Here is an example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the timecourse log fold change
computation.

```{r fig.height=4, fig.width=4}
pvalue = qvalues[, "M-K"]
lfc_timecourse = log_fold_change_timecourse[, "M-K"]
names(lfc_timecourse) = row.names(log_fold_change_timecourse)

plot(lfc_timecourse, -log10(pvalue), pch=20, main="Volcano plot",
     xlim=c(-2.5, 2))
```

@@

And here we visualize the genes with the largest absolute timecourse log-fold
change.

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_timecourse),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], splines_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

# convert gene names
genes = getBM(attributes=c("ensembl_gene_id", "entrezgene_id"),
		filters="refseq_mrna", values=genes,
		mart=ensembl)["entrezgene_id"]
genes = as.vector(unlist(genes))
pathways = find_enriched_pathway(
    genes, species="mmu",
    download_latest=FALSE)$stastic
